---
title: "2nd half of project"
author: "Juliet Alexandra"
date: "2025-07-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r, warning=FALSE}
suppressPackageStartupMessages({
library(dplyr)
library(tidyr)
library(fGarch)
library(gridExtra)
library(reshape2)
library(ggplot2)
library(ggpmisc)
library(tseries)
library(nortest)
library(zoo)
library(car)
library(lubridate)
library(purrr)
library(caret)
library(FinTS)
library(xts)
library(rugarch)
library(tibble)
library(forecast)})
```

Ensuring reproducibility of results:
```{r}
set.seed(48)
```

# Loading models and datasets
```{r}
ARIMA_model1 <- readRDS("Models/arima_model.rds")
ARIMAX_model2 <- readRDS("Models/arima_with_xreg.rds")
ARIMAX_model1 <- readRDS("Models/arimax_model.rds")
Energy_generation_model <- readRDS("Models/Energy_generation_model.rds")
GARCH_model <- readRDS("Models/garch_model.rds")
SARIMA_model1 <- readRDS("Models/sarima_model.rds")
STL_ARIMA_model1 <- readRDS("Models/stl_arima_model.rds")

```

```{r}
time_series_data <- read.csv("Saved_Datasets/timeseries.csv")
Training_set <- read.csv("Saved_Datasets/Training_set.csv")

Val_time_series <- read.csv("Saved_Datasets/Val_time_series.csv")
Val_set <- read.csv("Saved_Datasets/Val_set.csv")

xreg <- read.csv("Saved_Datasets/xreg.csv")
xreg_future <- read.csv("Saved_Datasets/xreg_future.csv")

xreg <- as.matrix(xreg)
xreg_future <- as.matrix(xreg_future)
```

```{r}
time_series_data <- ts(time_series_data$Value, start = c(2017, 1), frequency = 48)
Val_time_series <- ts(Val_time_series$Value, start = c(2018, 1), frequency = 48)
First_order_diff <- diff(time_series_data)
```

# Combining the training and test sets
I will be training the following models on both the 2017 and 2018 energy prices (my training set and validation set), then I will forecast the 2019 energy prices, and evaluating the models performance based on the actual 2019 energy prices (test set).

```{r}
# Combining train and validation sets
train_val <- bind_rows(as_tibble(time_series_data), as_tibble(Val_time_series))
train_val <- ts(train_val, start = c(2017, 1), frequency = 48)

# box-cox transformation
lambda <- BoxCox.lambda(train_val)
ts_bc <- BoxCox(train_val, lambda)

# First order differencing:
train_val_diff <- diff(train_val)

```

# Refitting models on combined dataset and forecasting:
### ARIMA
```{r}
ARIMA_model_Final_411 <- Arima(train_val, order = c(4,1,1))
```

I'm going to forecast 48 trading periods for 12 months
```{r}
forecast_ARIMA_411 <- forecast(ARIMA_model_Final_411, h = 17520)
forecast_ARIMA_411 <- ts(forecast_ARIMA_411$mean, start = c(2019, 1), frequency = 48)
```

```{r}
ARIMA_model_Final_312 <- Arima(train_val, order = c(3,1,2))
```

I'm going to forecast 48 trading periods for 12 months
```{r}
forecast_ARIMA_312 <- forecast(ARIMA_model_Final_312, h = 17520)
forecast_ARIMA_312 <- ts(forecast_ARIMA_312$mean, start = c(2019, 1), frequency = 48)
```

### SARIMA
```{r}
SARIMA_model_Final <- Arima(train_val, order = c(3,1,1),
                           seasonal = list(order = c(0, 0, 1), period = 48))
```

I'm going to forecast 48 trading periods for 12 months
```{r}
forecast_SARIMA_1 <- forecast(SARIMA_model_Final, h = 17520)
forecast_SARIMA <- ts(forecast_SARIMA_1$mean, start = c(2019, 1), frequency = 48)
```

### STL-ARIMA
```{r}
STL_ARIMA_FINAL <- stlm(train_val, s.window = "periodic", robust = TRUE,
                         modelfunction = function(x) Arima(x, order = c(5, 1, 2)))
```

I'm going to forecast 48 trading periods for 12 months
```{r}
forecast_STL_ARIMA_1 <- forecast(STL_ARIMA_FINAL, h = 17520)
forecast_STL_ARIMA <- ts(forecast_STL_ARIMA_1$mean, start = c(2019, 1), frequency = 48)
```

# Loading test data:
```{r}
csv_folder <- "2019"
csv_files <- list.files(path = csv_folder, pattern = "*.csv", full.names = TRUE)

test_data <- csv_files %>% lapply(read.csv) %>% bind_rows()
test_data <- test_data[test_data$PointOfConnection == "ABY0111", ]
test_data <- test_data[order(test_data$TradingDate, test_data$TradingPeriod), ]
test_data$TradingDate <- as.Date(test_data$TradingDate)

```

Because I removed daylight savings effected days from my training set then I am also going to remove these days from my test set. Daylight savings for 2019 starts on 2019-04-07

```{r}
ErrorIndices2 <- which(test_data$TradingPeriod > 48)
test_data <- test_data[-ErrorIndices2, ]
```

```{r}
test_data %>% mutate(test_data = as.Date(TradingDate)) %>% count(TradingDate) %>% filter(n < 48)
```

I'm going to use linear interpolation to fill in the 2 trading period gap for 2019-09-29
```{r}
# finding the row just before period 47 on 2019-09-29
i_prev <- which(test_data$TradingDate == as.Date("2019-09-29") &
                test_data$TradingPeriod == 46)

# Computing the two interpolated values: 
# Extracting the two known prices
price_prev <- test_data$DollarsPerMegawattHour[i_prev]
price_next <- test_data$DollarsPerMegawattHour[i_prev + 1]

# runing approx() over the X = {46,49} → Y = {prev,next}, get Y at Xout = {47,48}
interp <- approx(
  x    = c(46, 49),
  y    = c(price_prev, price_next),
  xout = c(47, 48),
  method = "linear"
)

# interp$x == c(47,48); interp$y == interpolated prices
interp
```

```{r}
new_rows <- tibble(
  TradingDate             = as.Date("2019-09-29"),
  TradingPeriod           = interp$x,
  PointOfConnection       = "ABY0111",
  DollarsPerMegawattHour  = interp$y
)

# bind back and re-sort
test_data <- bind_rows(test_data, new_rows) %>%
  arrange(TradingDate, TradingPeriod)

# view the gap now filled
filter(test_data,
       TradingDate == as.Date("2019-09-29"),
       TradingPeriod %in% 46:49)
```

Making it a time series object:
```{r}
price_vector2 <- as.numeric(test_data$DollarsPerMegawattHour)
Test_time_series <- ts(price_vector2, start = c(2019, 1), frequency = 48)
```

Double checking the start date, frequency, and length of my time series object:
```{r}
start(Test_time_series)
frequency(Test_time_series)
length(Test_time_series)
```

# Overall Model Comparison
```{r}
ARIMA_312_acc <- accuracy(forecast_ARIMA_312, Test_time_series)
ARIMA_411_acc <- accuracy(forecast_ARIMA_411, Test_time_series)
SARIMA_acc <- accuracy(forecast_SARIMA, Test_time_series)
STL_ARIMA_acc <- accuracy(forecast_STL_ARIMA, Test_time_series)
```

```{r}
acc_list <- list(ARIMA312 = ARIMA_312_acc, ARIMA411 = ARIMA_411_acc, SARIMA = SARIMA_acc, 
                  STL_ARIMA = STL_ARIMA_acc)

acc_df <- map_df(acc_list2, ~ as.data.frame(.x)["Test set", ], .id = "Model")
```

Manually calculating the MASE values:
```{r}
# MAE of forecast errors
mae_model <- mean(abs(Test_time_series - forecast_SARIMA))

# Naive forecast (random walk with no drift)
naive_forecast <- naive(train_val, h = length(Test_time_series))
naive_forecast <- ts(naive_forecast$mean, start = c(2019, 1), frequency = 48)
mae_naive <- mean(abs(Test_time_series - naive_forecast))

# MASE = MAE(model) / MAE(naive)
mase_SARIMA <- mae_model / mae_naive
mase_SARIMA
```

```{r}
mae_model <- mean(abs(Test_time_series - forecast_ARIMA_312))
naive_forecast <- naive(train_val, h = length(Test_time_series))
naive_forecast <- ts(naive_forecast$mean, start = c(2019, 1), frequency = 48)
mae_naive <- mean(abs(Test_time_series - naive_forecast))
mase_ARIMA3 <- mae_model / mae_naive
mase_ARIMA3
```

```{r}
mae_model <- mean(abs(Test_time_series - forecast_ARIMA_411))
naive_forecast <- naive(train_val, h = length(Test_time_series))
naive_forecast <- ts(naive_forecast$mean, start = c(2019, 1), frequency = 48)
mae_naive <- mean(abs(Test_time_series - naive_forecast))
mase_ARIMA4 <- mae_model / mae_naive
mase_ARIMA4
```

```{r}
mae_model <- mean(abs(Test_time_series - forecast_STL_ARIMA))
naive_forecast <- naive(train_val, h = length(Test_time_series))
naive_forecast <- ts(naive_forecast$mean, start = c(2019, 1), frequency = 48)
mae_naive <- mean(abs(Test_time_series - naive_forecast))
mase_STL <- mae_model / mae_naive
mase_STL
```

```{r}
acc_df <- acc_df %>% mutate(MASE = c(mase_ARIMA3, mase_ARIMA4, mase_SARIMA, mase_STL)) 

acc_df %>% arrange(MASE)
```

The MASE for STL-ARIMA is 0.79 which means it outperforms a naive forecast (its errors are 21% smaller than a naive forecasts).


## Plot
```{r}
start_time     <- as.POSIXct("2019-01-01 00:00:00", tz = "UTC")
times_forecast <- seq(start_time, by = "30 min", length.out = 17520)

Test_time_series2 <- ts(test_data$DollarsPerMegawattHour, frequency = 48, start = c(2019, 1))
df_actual <- data.frame(Date = times_forecast, Value = as.numeric(Test_time_series2), Type = "Actual")

forecast_raw_ARIMA_1_plot <- ts(forecast_SARIMA_1, start = c(2019, 1), frequency = 48)

df_forecast <- data.frame(
  Date  = times_forecast,
  Mean  = as.numeric(forecast_raw_ARIMA_1_plot$mean),
  Lower = as.numeric(forecast_raw_ARIMA_1_plot$lower[,2]),  # 95% lower bound
  Upper = as.numeric(forecast_raw_ARIMA_1_plot$upper[,2]),  # 95% upper bound
  Type  = "Forecast"
)

ggplot() +
  geom_line(data = df_actual, aes(x = Date, y = Value, color = "Actual"), size = 0.55) +
  geom_line(data = df_forecast, aes(x = Date, y = Mean, color = "Forecast"), size = 1, alpha = 0.8, linetype = "dashed") +
  geom_ribbon(data = df_forecast, aes(x = Date, ymin = Lower, ymax = Upper), fill = "lightblue", alpha = 0.3) +
  scale_color_manual(values = c("Actual" = "firebrick1", "Forecast" = "steelblue3")) +
  scale_x_datetime(date_breaks = "1 month",  date_labels = "%B",  expand = c(0, 0), guide = guide_axis(n.dodge = 2)) +
  labs(title = paste("SArima", " Forecast vs Actual Prices for ", 2019, sep = ""),
       x = "Month", y = "Electricity Price", color = "Legend") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# References and Citations

Electricity Authority. (n.d.). Final energy prices by month [Dataset]. EMI – Electricity Market Information. Retrieved between July 11 and July 15, 2025, from
https://www.emi.ea.govt.nz/Wholesale/Datasets/DispatchAndPricing/FinalEnergyPrices/ByMonth

Electricity Authority. (n.d.). Generation output by plant [Dataset]. EMI – Electricity Market Information. Retrieved between July 18 and July 20, 2025, from
https://www.emi.ea.govt.nz/Wholesale/Datasets/Generation/Generation_MD
